---
title: " A Statistical Analysis: Heart Disease Predection"
author: "Group Members: TY Punleu, Song Kimleangchhay Khy Pichsereyvathanak, Ouk Vatana, Srey Sitharath"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
  html_document:
    toc: true
    code_folding: hide
---

```{r setup, include=FALSE}
# This chunk sets global options for the notebook.
# It ensures all necessary packages are loaded and controls how code/output is displayed.
knitr::opts_chunk$set(
  echo = TRUE,      # Show the code in the final report
  warning = FALSE,  # Hide warnings
  message = FALSE,  # Hide messages
  fig.width = 7,    # Default width for figures
  fig.height = 5,   # Default height for figures
  cache = FALSE     # Set to TRUE only if processing time is long
)
```

```{r}
# TODO - Load essential libraries

```

# 1.Introduction

## 1.1. Problem Statement

Cardiovascular diseases remain the leading cause of death globally, including in Cambodia and Southeast Asia. Early identification of high-risk individuals using simple clinical measurements (age, cholesterol, blood pressure, etc.) can enable timely intervention and save lives. This project applies the full statistical workflow to the classic UCI Heart Disease dataset to uncover which factors most strongly predict the presence of heart disease and to build an interpretable prediction model. 
## 1.2. Research Objectives 
To address these problems, Our Objectives aim to identify the factors that increase the risk and help support early diagnosis and prevention by using statistical methods 
• To analyze if demographic (age or gender) influence heart disease 
• Find out which clinical features such as chest paint type, blood pressure, blood sugar, heart rate influence heart disease with strong affect 

## 1.3. Research Questions 
• Does age or gender influence heart disease the most? 
• Which clinical features that most prone to affect to the heart disease? 

# 2. Dataset 
## Numerical Variables

| Column   | Description                                          |
|----------|------------------------------------------------------|
| id       | Unique id for each patient                           |
| age      | Age of the patient in years                          |
| origin   | Place of study                                       |
| trestbps | Resting blood pressure                               |
| chol     | Serum cholesterol in mg/dl                           |
| thalch   | Maximum heart rate achieved                          |
| oldpeak  | ST depression induced by exercises relative to rest  |
| slope    | The slope of the peak exercise ST segment            |
| ca       | Number of major vessels (0–3) colored by fluoroscopy |
| num      | The predicted attribute                              |

## Categorical Variables

| Column | Description |
|-------------|-----------------------------------------------------------|
| sex | Male/Female |
| cp | Chest pain type (typical angina, atypical angina, non-anginal, asymptomatic) |
| fbs | Fasting blood (True/False) |
| restecg | Resting electrocardiographic results (normal, stt abnormality, lv hypertrophy) |
| exang | Exercise-induced angina (True/False) |
| ca | Number of major vessels (0–3) colored by fluoroscopy |
| thal | normal, fixed defect; reversible defect |

# 3. Methodology and Data Preprocessing

## 3.1. Data Acquisition and Cleaning

## 3.2. Descriptive Statistic

## 3.3. Inferential Statistic

## 3.4. Correlation and Regression Analysis

## 3.5. Intepretation and Recommendation
